---
title: "*Enterococcus faecalis* OG1RF Mutant Growth in Saturated Fatty Acids"
output: md_document
---

# Introduction

*Enterococcus faecalis* is a gram-positive human gut bacterium which is a pathogen of interest as it is often a causative factor of a large number and variety of opportunistic infections, including gastroenteritis, urinary tract infections, and bacteremia in immunocompromised patients. In previous research, it has been shown to uptake exogenous fatty acids, such as those often found in bile, into its membrane. However; the uptake of particular fatty acids, especially myristic acid and palmitic acid, occurs at some cost to the bacterium, often causing slowed growth and cell deformation. Mutant substrains of *E. faecalis* were previously generated by growing OG1RF in media supplemented with either myristic acid or palmitic acid dissolved in ethanol. Control substrains were also generated by growing OG1RF in media supplemented with ethanol. Here, we characterize the growth of these control and mutant strains of bacteria in several saturated fatty acids.

# Methods

Cultures of *E. faecalis* OG1RF, substrains EtOH#1 (control), MA#1, and PA#1 (mutants) were grown at 37&deg;C in 5 mL BHI (Brain heart infusion) overnight. These were diluted to optical densities of 0.01 at 600 nm in 25 mL of BHI and were supplemented with ethanol and each of the following fatty acids:

* Myristic acid (C14:0) at 5 &mu;g/mL
* Palmitic acid (C16:0) at 5 &mu;g/mL
* Stearic acid (C18:0) at 20 &mu;g/mL
* Lauric acid (C12:0) at 20 &mu;g/mL

The supplemented bacteria were grown at 37&deg;C and the optical density at 600 nm was measured every 30 min. The optical density was then plotted on a log scale against the elapsed time.

# Results
```{r}
####################Plots for growth of EtOH1, MA#1 and PA#1 in fatty acids#####################
library(ggplot2) # load ggplot2 library
data <- read.csv("../data/Sep15_2016_Growth_EtOH1_MA1_PA1_Strains.csv") # read in growth data
####Generate plot showing growth of EtOH#1 in the different saturated fatty acids and ethanol####
ggplot(data, aes(x=Time, y=value, color = Legend)) + 
  geom_line(aes(y=EtOH1_in_Ethanol_OD600, col = "EtOH#1 in Ethanol")) +
  geom_point(aes(y=EtOH1_in_Ethanol_OD600, col = "EtOH#1 in Ethanol")) +
  geom_line(aes(y=EtOH1_in_myristic_OD600, col = "EtOH#1 in Myristic Acid")) +
  geom_point(aes(y=EtOH1_in_myristic_OD600, col = "EtOH#1 in Myristic Acid")) +
  geom_line(aes(y=EtOH1_in_palmitic_OD600, col = "EtOH#1 in Palmitic Acid")) +
  geom_point(aes(y=EtOH1_in_palmitic_OD600, col = "EtOH#1 in Palmitic Acid")) +
  geom_line(aes(y=EtOH1_in_stearic_OD600, col = "EtOH#1 in Stearic Acid")) +
  geom_point(aes(y=EtOH1_in_stearic_OD600, col = "EtOH#1 in Stearic Acid")) +
  geom_line(aes(y=ETOH1_in_lauric_OD600, col = "EtOH#1 in Lauric Acid")) +
  geom_point(aes(y=ETOH1_in_lauric_OD600, col = "EtOH#1 in Lauric Acid")) +
  scale_y_log10() +
  ylab(expression(paste("Log ", OD[600], sep="")))+
  xlab("Time (min)")
```

The EtOH#1 substrain showed growth in the supplements which would be typically expected of wild-type *E. faecalis* OG1RF. Ethanol, stearic acid, and lauric acid affected the growth minimally, and both myristic and palmitic acid caused the optical density to level off at values between 0.2 and 0.3. The 24-hour time point showed that the optical density increased to just under 1.0, possibly indicating a takeover of the culture by a new mutant substrain after its prolonged exposure to myristic acid or palmitic acid.

```{r}
####Generate plot showing growth of MA#1 in the different saturated fatty acids and ethanol####
ggplot(data, aes(x=Time, y=value, color = Legend)) + 
  geom_line(aes(y=MA1_in_ethanol_OD600, col = "MA#1 in Ethanol")) +
  geom_point(aes(y=MA1_in_ethanol_OD600, col = "MA#1 in Ethanol")) +
  geom_line(aes(y=MA1_in_myristic_OD600, col = "MA#1 in Myristic Acid")) +
  geom_point(aes(y=MA1_in_myristic_OD600, col = "MA#1 in Myristic Acid")) +
  geom_line(aes(y=MA1_in_palmitic_OD600, col = "MA#1 in Palmitic Acid")) +
  geom_point(aes(y=MA1_in_palmitic_OD600, col = "MA#1 in Palmitic Acid")) +
  geom_line(aes(y=MA1_in_stearic_OD600, col = "MA#1 in Stearic Acid")) +
  geom_point(aes(y=MA1_in_stearic_OD600, col = "MA#1 in Stearic Acid")) +
  geom_line(aes(y=MA1_in_lauric_OD600, col = "MA#1 in Lauric Acid")) +
  geom_point(aes(y=MA1_in_lauric_OD600, col = "MA#1 in Lauric Acid")) +
  scale_y_log10() +
  ylab(expression(paste("Log ", OD[600], sep="")))+
  xlab("Time (min)")
```

The MA#1 substrain showed incredibly different growth characteristics in the fatty acids, with an especially strong difference in its growth when supplemented with myristic and palmitic acid. Instead of leveling off in the mid-log stage of growth as observed in EtOH#1, MA#1 appeared to be completely unfazed by supplementation with myristic and palmitic acid, and grew in it as if it were completely unsupplemented. Lauric acid appeared to have a slight effect on the growth of MA#1, as it leveled off at a slightly lower optical density than observed in growth supplemented with ethanol or the other saturated fatty acids.

```{r}
####Generate plot showing growth of PA#1 in the different saturated fatty acids and ethanol####
ggplot(data, aes(x=Time, y=value, color = Legend)) + 
  geom_line(aes(y=PA1_in_ethanol_OD600, col = "PA#1 in Ethanol")) +
  geom_point(aes(y=PA1_in_ethanol_OD600, col = "PA#1 in Ethanol")) +
  geom_line(aes(y=PA1_in_myristic_OD600, col = "PA#1 in Myristic Acid")) +
  geom_point(aes(y=PA1_in_myristic_OD600, col = "PA#1 in Myristic Acid")) +
  geom_line(aes(y=PA1_in_palmitic_OD600, col = "PA#1 in Palmitic Acid")) +
  geom_point(aes(y=PA1_in_palmitic_OD600, col = "PA#1 in Palmitic Acid")) +
  geom_line(aes(y=PA1_in_stearic_OD600, col = "PA#1 in Stearic Acid")) +
  geom_point(aes(y=PA1_in_stearic_OD600, col = "PA#1 in Stearic Acid")) +
  geom_line(aes(y=PA1_in_lauric_OD600, col = "PA#1 in Lauric Acid")) +
  geom_point(aes(y=PA1_in_lauric_OD600, col = "PA#1 in Lauric Acid")) +
  scale_y_log10() +
  ylab(expression(paste("Log ", OD[600], sep="")))+
  xlab("Time (min)")
```

PA#1 grew differently than either EtOH#1 or MA#1. Notably, palmitic acid appeared to have a less detrimental effect on the growth of PA#1 than it had on the growth of EtOH#1, although myristic still appeared to have a detrimental effect, if not as pronounced as in EtOH#1.

# Discussion

The differences of growth between the control strains and mutant strains were remarkable, especially when supplemented with myristic and palmitic acid. Myristic and palmitic acid appeared to have a less detrimental effect on the growth of MA#1 and PA#1 than on EtOH#1, although there were apparent differences between the MA#1 and PA#1 strains as well. One possible hypothesis which might explain this phenomenon is that certain mutations may have arisen in the mutant strains' DNA which allowed them to better handle growth in myristic and palmitic acid. 